---
title: "Lab 9 Intro"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("jpeg")
```

## Seating
1. If you already have a group, it will be helpful to start sitting next to each other.
2. If you don't have a group, hang tight we are working on it.
3. Please setup your computer as normal for today.

```{r}
install.packages("here")
```
```{r}
library(here)
library(janitor)
library(tidyverse)
```

## Warm-up
1. In the data folder there is an epidemiology data set on an outbreak of malaria.
```{r}
malaria <-readr::read_csv("data/malaria.csv")
malaria
```

2. `rdt` refers to rapid diagnostic test and they are identified here by age group.


3. Make the data tidy and store them as a new object.
```{r}
malaria <-read_csv(here("lab9", "data", "malaria.csv"))%>% 
  clean_names()

```

```{r}
malaria_k <-
malaria %>% 
  pivot_longer(cols=starts_with("malaria_rdt"),
               names_to = "age_class",
              values_to= "cases") %>% 
  select(newid,district,submitted_date,malaria_tot)
malaria_k

```
malaria_k<-

select(District,submitted_date,malaria_tot)
4. Which district had the highest *total* number of cases on July 30, 2020?
```{r}
malaria_k %>% 
  filter(date == "2020-07-30") %>%
  group_by(District) %>% 
  summarize(tot_cases=sum(cases, na.rm=T)) %>% 
  

```


** check the warmup **





