---
title: "Lab 3 Warmup"
output: html_document
date: "2023-01-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Warm-ups
1. Build a vector that includes the following height measurements for five plants:
plant 1 30.7, plant 2 37.6, plant 3 28.4, plant 4 NA, plant 5 33.2
## plant Height
```{r}
plant_height<- c(30.7,37.6,28.4,NA,33.5)
```

2. Build another vector that includes the following mass measurements:
plant 1 4, plant 2 5.2, plant 3 3.7, plant 4 NA, plant 5 4.6
```{r}
plant_mass<- c(4,5.2, 3.7,NA,4.6)
```

vector
```{r}
plant_experiment<-c(plant_height,plant_mass)
plant_experiment
```

3. Assemble these vectors into a labeled data matrix with two columns
```{r}
plant_experiment_matrix<-matrix(plant_experiment,nrow=5,byrow=F)
plant_experiment_matrix
```

## Name colums and rows

```{r}
colnames()<-(plant_experiment_matrix)<- measure
rownames<-(plant_experiment_matrix)<-samples
```
###
4. Calculate the mean for height and mass and add them to the data matrix








